---
title: "Spliced peptides in colon cancer (interim results)"
author: "Jason Steen"
date: "`r Sys.Date()`"
output: html_document
---

### Background

Given a list of peptides and some fastq files, identify how often each peptide is present in each sample

Two data archives are given, 

 * "GSA" https://ngdc.cncb.ac.cn/gsa/browse/CRA001257 
 * "GEO" https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE163067  

**Step 1:** download data from GSA

List of accessions
```
ftp://download.big.ac.cn/gsa/CRA001257/CRR038730/CRR038730.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038729/CRR038729.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038728/CRR038728.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038727/CRR038727.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038726/CRR038726.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038725/CRR038725.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038724/CRR038724.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038723/CRR038723.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038722/CRR038722.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038721/CRR038721.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038720/CRR038720.fq.gz
ftp://download.big.ac.cn/gsa/CRA001257/CRR038719/CRR038719.fq.gz
```

**Step 2:** translate each read into each of 6 reading frames and search them for the peptides in the list patterns.
I deleted the * that appears to signify either a L or an I from the peptides that contained it. We can interrogate this further by context of each peptide.
```{python, python.reticulate=FALSE, eval=FALSE}
#!/usr/bin/python3

patterns = [
"ALRPSTSSL",
"WLMPVIPAL",
"SADDIIIYL",
"HLLQEELIL",
"ILPKVIYRF",
"GMYLKIIIT",
"QLMALQQGL",
"SLTKSLPAL",
"VLLSNYASL",
"YLDELLQVL",
"YVDDLYVVL",
"ELMKLLEEL",
"HLYNLLLL",
"KVPVGPVLV",
"LLDQKDQLV",
"LLLSNYASL",
"IDDKGTIKL",
"LIDDKGTIKL",
"LLIDDKGTIKL",
"NLLIDDKGTIKL",
"ALIDDKGTIKL",
"HLIDDKGTIKL",
"LAIDDKGTLKL",
"LGIDDKGTLKL",
"LSIDDKGTLKL",
"SLIDDKGTIKL",
"TLIDDKGTIKL",
"GMYLKIIIT",
"HLLQEELIL",
"LLLPAEVAIVHV",
"SADDIIIYL",
"FLFQDFLRQA",
"FLIDDKGTIKL",
"QYQQLLRSL",
"SPIAERSMM"
]

with gzip.open(input_file, "rt") as file:
    output = open(output_file, "a")
    reads_to_keep = []
    for record in SeqIO.parse(file, "fastq"):
        sequence = record.seq
        sequence_rev = sequence.reverse_complement()
        for frame_start in range(3):
            frame = translate(sequence[frame_start:])
            rev_frame = translate(sequence_rev[frame_start:])
            for pattern in patterns:
                if pattern in frame:
                    record_out = str("found " + pattern + " in " + sequence + " " + frame + "\n")
                    reads_to_keep.append(record)
                    output.write(record_out)            
            for pattern in patterns:
                if pattern in rev_frame:
                    record_out2 = str("found " + pattern + " in " + sequence_rev + " " + rev_frame + "\n")   
                    reads_to_keep.append(record)
                    output.write(record_out2)
    SeqIO.write(reads_to_keep, output_fastq, "fastq")


```

**Step 3:** add a quick header to each file so its easier to combine in R

```{bash, eval=FALSE}
#first subset the three interesting columns
for i in *.txt; do awk 'BEGIN{FS=" ";OFS="\t"}{print $2, $4, $5}' $i > $i.cut; done
#then add the header to each file, header.txt contains a single line "peptide read  context"
for i in *.cut; do cat header.txt  $i > $i.head; done
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
require(fs)
require(tidyr)
require(purrr)
require(readr)
require(knitr)
```

**Step 4:** take GSA data and combine them into a single dataframe, then display the table
```{r GSA_data, message=FALSE, warning=FALSE}

Files <- dir_ls("./", glob="*.head")
List <- map(Files, read_delim) 
Bind <- bind_rows(List, .id = 'Sample', ) %>% separate(Sample, into=c("Sample", NA, NA, NA), sep="[.]", remove=TRUE)
a <- Bind %>% group_by(Sample, peptide) %>% summarise(n=n()) %>% pivot_wider(names_from=Sample, values_from=n)
kable(a, caption = "Counts of each peptide per sample from GSA data")
```


**Step 5:** take SRA data and and combine them into a single dataframe, then display the table
```{r SRA_data, message=FALSE, warning=FALSE}

Files_SRA <- dir_ls("./SRA_data/", glob="*.head")
List_SRA <- map(Files_SRA, read_delim) 
Bind_SRA <- bind_rows(List_SRA, .id = 'Sample', ) %>% separate(Sample, into=c(NA, NA, NA, NA, "Sample", NA, NA, NA, NA), sep="[/_.]", remove=TRUE)
b <- Bind_SRA %>% group_by(Sample, peptide) %>% summarise(n=n()) %>% pivot_wider(names_from=Sample, values_from=n)
kable(b, caption = "Counts of each peptide per sample from SRA data")
```



#### Limitations and caveats
 * this is interim data for evaluation of the method and should not be construed as a final result
 * This counts raw data from fastq files.  we have not evaluated quality of basecalls in these reads
 * peptides that are subsets of each other are likely counted together (ie, counts for "IDDKGTIKL" could well belong to either of the longer "LIDDKGTIKL" or "LLIDDKGTIKL" peptides.
 * peptides that are not entirely present on a read will not be counted. we have made no attempts to assemble the reads into longer sequences. 


